Gravitational wave sources in our Galactic backyard: LISA predictions for BH and NS binaries

### 2021-06-18 paper/tex/sections/discussion.tex
finish distinguishing

\subsubsection{Distinguishing from WDWD population}
The population of Galactic WDWDs detectable with LISA will be several orders of magnitude larger than the more massive DCOs on which we focus in this paper \citep[e.g.][]{Korol+2017}. It is therefore imperative that we consider how to distinguish from this much more numerous population of sources.

The easiest way to check whether a source is a WDWD is to check its chirp mass. The mass of a non-rotating white dwarf cannot be larger than the Chandrasekhar limit of $1.44 \unit{M_\odot}$ \citep{Chandrasekhar+1931}, so we can take the maximum chirp mass of a WDWD to be $\sim 1.25 \unit{M_{\odot}}$. Therefore, any DCO with a chirp mass that satisfies $\mathcal{M}_c > 1.25 \unit{M_{\odot}} + 2 \Delta \mathcal{M}_c$ must not be a WDWD. We find that 17\%(23\%) of BHBHs, 20\%(24\%) of BHNSs and 2.4\%(2.5\%) of NSNSs for a 4(10)-year LISA mission satisfy this condition.

\todo{Another bit about the fact that eccentric DCOs that are resolved within the disc can be quite confidently confirmed as not WDWDs}

@@

\todo{}

>>>>

\subsubsection{Distinguishing from WDWD population}\label{sec:WDWD_distinguish}
The population of Galactic WDWDs detectable with LISA will be several orders of magnitude larger than the more massive DCOs on which we focus in this paper \citep[e.g.][]{Korol+2017}. It is therefore imperative that we consider how to distinguish NS and BH binaries from this much more numerous population of sources.

The simplest way to check whether a source is a WDWD is to check its chirp mass. The mass of a non-rotating white dwarf cannot be larger than the Chandrasekhar limit of $1.44 \unit{M_\odot}$ \citep{Chandrasekhar+1931}, so we can take the maximum chirp mass of a WDWD to be $\sim 1.25 \unit{M_{\odot}}$. Therefore, any DCO with a chirp mass that satisfies $\mathcal{M}_c > 1.25 \unit{M_{\odot}} + 2 \Delta \mathcal{M}_c$ must not be a WDWD. We find that for a 4(10)-year LISA mission, 17(23)\% of BHBHs, 20(24)\% of BHNSs and 2.4(2.5)\% of NSNSs satisfy this condition. As one would expect, this method is not particularly effective for NSNSs since their average chirp mass, $1.17 \unit{M_\odot}$, is below the Chandrasekhar limit.

Another indicator that the source is not a WDWD would be that it is eccentric. WDWDs formed in the disc are thought to be formed through isolated binary formation and have little to no eccentricity \citep[e.g.][]{Nelemans+2001}. Therefore, if any system is detected with anything other than one detectable harmonic, this implies that the system is unlikely to be a WDWD. We find that for a 4(10)-year LISA mission, 55(61)\% of BHBHs, 27(28)\% of BHNSs and 65(69)\% of NSNSs are detected with multiple harmonics. Both the absolute percentage and the improvement with an extended LISA mission is lower for the BHNSs as we find that these DCOs are less eccentric on average (see Fig.~\ref{fig:fiducial_pdf_distributions}).

However, we should also consider that eccentric WDWDs \textit{could} be formed through dynamical formation in Milky Way globular clusters \citep[e.g.][]{Willems+2007, Kremer+2018}. This means that it would not be valid to assume that eccentric binaries are not WDWDs unless they are detected in the Galactic plane. Therefore, we can use the sky localisation, scale height of the disc and distance to the source to estimate what fraction of eccentric sources can be localised to the Galactic plane. This condition can be written as $\sigma_\theta < \arcsin(z_{\rm plane} / D_L)$ or $D_L < z_{\rm plane}$, where we set the height of the Galactic plane, $z_{\rm plane}=0.95 \unit{kpc}$, to the scale height of the high-$\alpha$ disc. We apply this condition to find that the fraction of sources that are eccentric \textit{and} localised within the disc for a 4(10)-year LISA mission are 40(32)\% for BHBHs, 24(19)\% for BHNSs and 59(51)\% for NSNSs\footnote{Note that although the fractions are smaller for the 10 year mission, the \textit{absolute} number of detections is still greater. The fraction decreases because a 10 year mission detects more `marginal' sources that are just on the cusp of the detection threshold and these sources have the worst sky localisation and thus cannot be confirmed to lie within the Galactic plane.}.

Overall, combining these methods we find that for a 4(10)-year mission, LISA will detect at least 15(22) BHBHs, 12(18) BHNSs and 19(27) NSNSs that are distinguishable from the WDWD population.

@@

The problem of separating the BHBH, BHNS and NSNS population can be more difficult. We can follow a similar method to the WDWDs (see Sec.~\ref{sec:WDWD_distinguish}) by applying our knowledge of the maximum mass of a neutron star. Following our fiducial assumption, we can take the maximum mass of a neutron star as $2.5 \unit{M_{\odot}}$ and thus the maximum chirp mass that a system can attain without one of the components being a black hole is $\mathcal{M}_{c} = 2.2 \unit{M_\odot}$. For a 4(10)-year LISA mission, the fraction of systems that are above or below this limit by more than $2 \Delta \mathcal{M}_c$ is 15(20)\% for BHBHs, 12(14)\% for BHNSs and 34(40)\% of NSNSs, which in terms of absolute detections is 5(11) for BHBHs, 4(8) for BHNSs and 11(21) for NSNSs. For separating the BHBH and BHNS population one could consider the range of mass ratios that could result in the chirp mass and assume that those that are likely close to unity must be BHBHs.

Another possible solution would be the existence of electromagnetic counterparts to the gravitational wave signal. In the next section (Sec.~\ref{sec:pulsar_matching}) we consider the possibility of detecting a pulsar within a BHNS or NSNS system. This could be used to identify the type of the source, however it is unlikely that a large fraction of the population will contain pulsars that are beaming towards the Earth.

One could also consider using the eccentric or orbital frequency to separate the populations since the distributions are reasonable different for each DCO type (see Fig.~\ref{fig:fiducial_pdf_distributions}). This method would also pose a challenge as it would likely only favour one DCO type as the source of the signal rather than provide strong evidence as the chirp mass could.

### 2021-06-18 paper/tex/sections/results.tex
finish distinguishing

In this section we present our main results for the detectable LISA DCO population. We find that, for our fiducial model, a four year LISA mission will detect \confinv{30.6}{20.5}{51.1} BHBHs, \confinv{26.4}{17.3}{48.5} BHNSs and \confinv{7.1}{4.8}{10.8} NSNSs where the error bars represent the 90\% confidence interval. We first show the distribution of the sources on the sensitivity curve in Section~\ref{sec:dcos_on_sc}, before exploring the variations in the detection rate over different physics variations in Section~\ref{sec:detection_rate_analysis} and analysing the parameter distributions for detectable sources in the fiducial model in Section~\ref{sec:fiducial_distributions}.

@@

\subsection{Detection rates}\label{sec:detection_rate_analysis}
\todo{I will be updating this section once the other physics variations are done running. There will be a couple of new models and the current ones will have better high Z resolution.}
We find that for our fiducial model, a four year LISA mission will detect \confinv{30.6}{20.5}{51.1} BHBHs, \confinv{26.4}{17.3}{48.5} BHNSs and \confinv{7.1}{4.8}{10.8} NSNSs, where the error bars represent the 90\% confidence interval. Increasing the LISA mission length to ten years changes the number of detections to \confinv{50.6}{29.2}{63.3}, \confinv{46.2}{25.7}{99.3} and \confinv{12.1}{6.9}{13.5} respectively. In Figure~\ref{fig:detection_rates}, we show the expected number of LISA detections for each model variation and discuss the prominent trends in the following sections. We show the rates and uncertainties plotted in this figure in Table~\ref{tab:detection_rates}.

\begin{figure*}[p]
    \centering
    \includegraphics[width=\textwidth]{3_dco_detections.pdf}
    \caption{The number of expected detections in the LISA mission for different DCO types and model variations. Error bars show the 50\% (solid) and 90\% (dotted) confidence intervals. The left axis and grid lines show the number of detections in a four year LISA mission and the right axis shows an approximation of the number of detections in a 10 year mission (we scale the axis by $\sqrt{T_{\rm obs}}$, see Table~\ref{tab:detection_rates} for exact rates). Each model is described in further detail in Table~\ref{tab:physics_variations} and details of the fiducial assumptions are in Section~\ref{sec:fiducial_physics}. \todo{subject to change with the updated/new models}}
    \label{fig:detection_rates}
\end{figure*}

@@

\begin{figure*}[htbp]
    \centering
    \includegraphics[width=\textwidth]{4_distribution_grid_4yr.pdf}
    \caption{Distributions for various parameters of the DCOs that are detectable in a 4 year LISA mission in our fiducial model. Each panel shows the distribution of a single parameter, where the colour denotes the DCO type. We also plot the 1- and 2-$\sigma$ uncertainties (obtained via bootstrapping) with the dark and light shaded areas respectively. The first two rows (excluding metallicity) use kernel density estimators to show the distributions. The dotted lines in the black hole mass panel show the individual distributions of the primary and secondary masses. The metallicity panel shows the distribution over the metallicity bins used in our population synthesis, which we show in the grid lines. The final three panels for the cumulative distribution functions for observables, normalised to the expected number of fiducial detections in a four year LISA mission. The dark shaded areas indicate regimes in which the quantity cannot be measured. The dotted lines in the angular resolution plot show the maximum angular resolution that can be covered by a single pointing of the labelled instrument. In Section~\ref{sec:fiducial_distributions} we discuss the features of the distributions.}
    \label{fig:fiducial_pdf_distributions}
\end{figure*}

@@

\subsection{Model variations}
\todo{This will be about how the shapes of the parameter distributions change for the different model variations}

>>>>

In this section we present our main results for the detectable LISA DCO population. We find that, for our fiducial model, a four 4-year LISA mission will detect \confinv{30.6}{20.5}{51.1} BHBHs, \confinv{26.4}{17.3}{48.5} BHNSs and \confinv{7.1}{4.8}{10.8} NSNSs where the error bars represent the 90\% confidence interval. We first show the distribution of the sources on the sensitivity curve in Section~\ref{sec:dcos_on_sc}, before exploring the variations in the detection rate over different physics variations in Section~\ref{sec:detection_rate_analysis} and analysing the parameter distributions for detectable sources in the fiducial model in Section~\ref{sec:fiducial_distributions}.

@@

\subsection{Detection rates}\label{sec:detection_rate_analysis}
\todo{I will be updating this section once the other physics variations are done running. There will be a couple of new models and the current ones will have better high Z resolution.}
We find that for our fiducial model, a 4-year LISA mission will detect \confinv{30.6}{20.5}{51.1} BHBHs, \confinv{26.4}{17.3}{48.5} BHNSs and \confinv{7.1}{4.8}{10.8} NSNSs, where the error bars represent the 90\% confidence interval. Increasing to a 10-year LISA mission length changes the number of detections to \confinv{50.6}{29.2}{63.3}, \confinv{46.2}{25.7}{99.3} and \confinv{12.1}{6.9}{13.5} respectively. In Figure~\ref{fig:detection_rates}, we show the expected number of LISA detections for each model variation and discuss the prominent trends in the following sections. We show the rates and uncertainties plotted in this figure in Table~\ref{tab:detection_rates}.

\begin{figure*}[p]
    \centering
    \includegraphics[width=\textwidth]{3_dco_detections.pdf}
    \caption{The number of expected detections in the LISA mission for different DCO types and model variations. Error bars show the 50\% (solid) and 90\% (dotted) confidence intervals. The left axis and grid lines show the number of detections in a 4-year LISA mission and the right axis shows an approximation of the number of detections in a 10-year mission (we scale the axis by $\sqrt{T_{\rm obs}}$, see Table~\ref{tab:detection_rates} for exact rates). Each model is described in further detail in Table~\ref{tab:physics_variations} and details of the fiducial assumptions are in Section~\ref{sec:fiducial_physics}. \todo{subject to change with the updated/new models}}
    \label{fig:detection_rates}
\end{figure*}

@@

\begin{figure*}[htbp]
    \centering
    \includegraphics[width=\textwidth]{4_distribution_grid_4yr.pdf}
    \caption{Distributions for various parameters of the DCOs that are detectable in a 4-year LISA mission in our fiducial model. Each panel shows the distribution of a single parameter, where the colour denotes the DCO type. We also plot the 1- and 2-$\sigma$ uncertainties (obtained via bootstrapping) with the dark and light shaded areas respectively. The first two rows (excluding metallicity) use kernel density estimators to show the distributions. The dotted lines in the black hole mass panel show the individual distributions of the primary and secondary masses. The metallicity panel shows the distribution over the metallicity bins used in our population synthesis, which we show in the grid lines. The final three panels for the cumulative distribution functions for observables, normalised to the expected number of fiducial detections in a 4-year LISA mission. The dark shaded areas indicate regimes in which the quantity cannot be measured. The dotted lines in the angular resolution plot show the maximum angular resolution that can be covered by a single pointing of the labelled instrument. In Section~\ref{sec:fiducial_distributions} we discuss the features of the distributions.}
    \label{fig:fiducial_pdf_distributions}
\end{figure*}

@@

\subsection{Model variations}
\todo{This will be about how the shapes of the parameter distributions change for the different model variations. I won't detail every variation but I'll point out anything that stands out and leave the rest in the appendix plots.}
